Machine Learning for Predictive Auto-Tuning

### 2011-12-13 paper/draft1.tex

Due to power consumption and heat dissipation concerns, scientific applications
have shifted from computing platforms where performance had been primarily
driven by rises in the clock frequency of a single ''heavy-weight'' processor
(with complex out-of-order control and cache structures) to a platform with
ever increasing numbers of ''light-weight'' cores. Interestingly, this shift is
now not only relevant to computational sciences but to the development of all
computer systems: from ubiquitous consumer-facing devices (e.g. smart phones) to
high-end computer farms for web-scale applications (e.g. social networks).

% HPC Landscape: large diversity of architectures, no consensus => push to flexibility => push to sw

Although the future lies in low-power multi-core hardware designs, the field
lacks meaningful consensus on exactly how the different subsystems (memory,
communication and computation) should be efficiently integrated, modelled and
programmed. These systems have exhibited varying degrees of memory hierarchy
and multi-threading complexity and, as a consequence, they have been
increasingly relying on flexible but low-level software-controlled cache
management and parallelism \cite{BerkeleyViewOfHPC} in order to better control
and understand the various trade-offs among performance, reliability, energy
efficiency, production costs, etc. This evolution has profoundly altered the
landscape of application development: programmers are now facing a wide
diversity of low-level architectural features to write high-performance
\emph{and} portable code.

% 1.1 Motivation
% Better tools: motivation (compilers/tools/libs can't keep up, it's too difficult)

In this rapidly evolving landscape, the construction of \emph{general}
development tools and libraries that \emph{fully utilize} system resources
remains a daunting task \cite{?}. Even within specialized architectures from
the \emph{same} vendor, like NVIDIA's Graphics Processing Units (GPUs) and the
Compute Unified Device Architecture (CUDA) \cite{CUDA}, many developers default
to massive amounts of manual labor to optimize CUDA code to specific input
domains. In addition, hand-tuning rarely generalizes well to new hardware
generations or different input domains, and it can also be error-prone or far
from optimal. One of the reason is that kernels can produce staggeringly large
optimization spaces \cite{K. Datta et al. Stencil computation optimization and
auto-tuning on state-of-the-art multicore architectures}. The problem is
further compounded by the fact that these spaces can be highly discontinuous
\cite{RyooHwuCGO2008}, difficult to explore and holding quasi-optimal solutions
at the edge of ''performance cliffs'' (in part due to hard resource
restrictions, like the register file size).

% COMMENT: NP, great setup! -JB

\section{Auto-Tuning}

% 1.2 related work
% auto-tuning is an answer

One of the strategies for addressing these challenges is to use automatic
methods called ''auto-tuning''. The literature on auto-tuning is extensive (see
surveys in \cite{XXX}. Two different approaches to auto-tuning have emerged:
model-driven optimization and empirical optimization.

The model-driven optimization approach uses analytical abstractions to model
the hardware architectures, the possible code transformations and their complex
interactions \cite{XXX}. Even though highly-accurate analytical models are
generally intractable to build, this approach has been quite successful in the
past at accelerating serial code \cite{XXX} with simplified but general
abstractions. However, large speed-ups for parallel code require more accurate
high-dimensional models and since this approach is bound by the quality and
scalability of its abstraction, it has been less suited for highly-specialized
% XXX: the bit above is confusing
kernels. This approach has been dominant in the compiler community \cite{XXX}.
As a result, it has generally been applied at compile-time where important
run-time characteristics such as input domains are missing. These drawbacks
render the model-driven optimization approach less attractive for
high-performance library developers.

% XXX: * empirical optimization: intro / advantages
% XXX: * empirical optimization: disadvantages
% XXX: * hybrid optimization (to counter disadvantages) ?
% XXX: * empirical optimization: w/ ML @ compile time (another way to “fix” empirical)
% XXX: * empirical optimization: ML @ runtime, especially for libs
% XXX: * empirical optimization: ML @ runtime for GPU libs why ? (w/ rtcg jit, rel. to pycuda, etc)

>>>>

Due to power consumption and heat dissipation concerns, scientific applications
have shifted from computing platforms where performance had been primarily
driven by rises in the clock frequency of a single ``heavy-weight'' processor
(with complex out-of-order control and cache structures) to a platform with
ever increasing numbers of ``light-weight'' cores. Interestingly, this shift is
now not only relevant to computational sciences but to the development of all
computer systems: from ubiquitous consumer-facing devices (e.g. phones) to
high-end computer farms for web-scale applications (e.g. social networks).

% HPC Landscape: large diversity of architectures,
% no consensus => push to flexibility => push to sw

Although the future lies in low-power multi-core hardware designs, the field
lacks meaningful consensus on exactly how the different subsystems (memory,
communication and computation) should be efficiently integrated, modeled and
programmed. These systems have exhibited varying degrees of memory hierarchy
and multi-threading complexity and, as a consequence, they have been
increasingly relying on flexible but low-level software-controlled cache
management and parallelism \citep{asanovic2006landscape} in order to better
control and understand the various trade-offs among performance, reliability,
energy efficiency, production costs, etc. This evolution has profoundly
altered the landscape of application development: programmers are now
facing a wide diversity of low-level architectural features to write
high-performance \emph{and} portable code.

\subsection{Motivation}

% Better tools: motivation (compilers/tools/libs can’t keep up, it’s too
% difficult)

In this rapidly evolving landscape, the construction of \emph{general}
development tools and libraries that \emph{fully utilize} system resources
remains a daunting task. Even within specialized architectures from the
\emph{same} vendor, like NVIDIA's Graphics Processing Units (GPUs) and the
Compute Unified Device Architecture (CUDA) \citep{nickolls2008scalable,
nvidia2011cuda}, many developers default to massive amounts of manual labor
to optimize CUDA code to specific input domains. In addition, hand-tuning
rarely generalizes well to new hardware generations or different input
domains, and it can also be error-prone or far from optimal. One of the
reason is that kernels can produce staggeringly large optimization spaces
\citep{datta2008stencil}. The problem is further compounded by the fact that
these spaces can be highly discontinuous \citep{ryoo2008program}, difficult
to explore and holding quasi-optimal solutions at the edge of ``performance
cliffs'' (in part due to hard resource restrictions, like the register file
size).

\subsection{Auto-Tuning}
% Related Work instead ??

% auto-tuning is an answer

One of the strategies for addressing these challenges is to use automatic
methods called ``auto-tuning''. Two major auto-tuning approaches have emerged
in the extensive literature covering the subject (see surveys in
\citep{vuduc2001statistical, demmel2005self, vuduc2005oski, williams2008auto,
datta2008stencil, cavazos2008intelligent, li2009note, park2011evaluation}):
analytical model-driven optimization and empirical optimization
\citep{yotov2003comparison}.

% model-driven optimization: intro

The model-driven optimization approach uses analytical abstractions to model
the hardware architectures, the possible code transformations and their complex
interactions. Even though highly-accurate analytical models are generally
intractable to build, this approach has been quite successful in the past at
accelerating serial code with simplified but general abstractions. However,
large speed-ups for parallel code require more accurate high-dimensional models
and since this approach is bound by the quality and scalability of its
abstraction, it has been less suited for highly-specialized kernels. This
approach has been dominant in the compiler community. As a result, it has
generally been applied at compile-time where important run-time characteristics
such as input domains are missing. These drawbacks render the model-driven
optimization approach less attractive for high-performance library developers.

% * empirical optimization: intro / advantages

The empirical optimization approach, on the contrary, seeks to find the best
performing code configuration by automatically generating many versions of a
parametrized kernel and benchmarking them on the actual hardware (possibly at
runtime, when contextual information about the hardware and software stack is
the richest). This method directly optimizes the metric(s) of interest and not
surrogates. One of its main advantage is that it allows any metric to be
optimized without loss of generality. It is indeed possible to formulate the
problem as a multi-objective optimization and minimize both speed \emph{and}
power consumption \citep{rahman2011automated}, a feat that renders the
analytical model-driven approach even more intractable. Due to its flexibility,
empirical auto-tuning has been successfully applied to build high-performance
domain-specific libraries including dense linear algebra
\citep{clint2001automated, bilmes1997optimizing}, sparse linear algebra
\citep{vuduc2005oski}, signal processing \citep{frigo1998fftw}, sorting
\citep{li2004dynamically}, general stencil operations \citep{kamil2010auto}, etc.

Even though this ``black-box'' approach does not necessarily try to build an
explicit model per se, it could, if care is taken, help to greatly accelerate
our understanding of the complex interactions between the set of optimizations,
their parameters and the hardware architectural features (XXX: that's what
we are trying to do w/ ML-based, decision trees, etc.).

% empirical optimization: disadvantages

The empirical approach is very sensible to the choice of instrumented
optimizations \emph{and} to the search method. The size of the search space is
so large that the current best empirical auto-tuners can only consider
highly-specialized functions with a limited set of code transformations and
compiler options, on very few input domains \citep{ganapathi2009case}. Although
searching for good code configurations in this highly-discontinuous space can
be made embarrassingly parallel, it remains a very difficult and prohibitively
expensive combinatorial optimization problem as many variants of the code must
generated, compiled, and benchmarked on specific input domains with meaningful
statistics (that may require multiple runs). Consequently, most proposed
methods prune the space with hard-coded heuristics that offer little
generalization guarantees. This has been the main drawback of the empirical
approach compared to the model-driven approach where good code configurations
can be directly derived from the analytical model.

% hybrid optimization: the bad way

To alleviate this major weakness, it seems intuitive to combine the two
approaches by first constraining the search space with an analytical model and
then exploring the reduced space empirically \citep{chen2005combining,
li2009note}. Unfortunately, this hybrid approach is still bound by the quality
of the analytical model, which remains hard to build by hand.

% hybrid optimization: the ML way (note that ML-based are in essence
% model-driven too just with an empirically learned model)

In this paper, we propose to \emph{learn the model}, empirically, (XXX:
we'll describe what this means in more details later in the paper, riiiighhht
???) instead of deriving it analytically. By learning the model, one can
hope to get ``the best of both approaches'' and capitalize on the large body of
literature from the machine learning community.

Although various statistical prediction techniques have been applied with
success at compile-time for general programs on various CPU architectures
\citep{monsifrot2002machine, stephenson2003meta, yotov2003comparison,
kulkarni2004fast, cooper2005acme, franke2005probabilistic,
hutter2006performance, cavazos2007rapidly, cavazos2008intelligent,
hartono2009annotation, park2011evaluation, fursin2008milepost}, there is, to
our knowledge, no fast predictive auto-tuning work that has tried to cover
simultaneously: (a) more recent multi-core architectures like GPUs
\citep{schaa2009exploring}, (b) high-performance domain-specific libraries
\citep{nukada2009auto, li2009note, kamil2010auto}, (c) at run-time
\citep{klockner2011pycuda, pinto+cox:2011gcg}, and (d) on \emph{many} input
domains \citep{liu2009cross, grauer2011optimizing}.

XXX WIIFY here ??? (what's in their for you) XXX in text, say what's in this paper for:
- compiler people
- libs people
- ml people
- tools people
- cv people

% auto-tuning is an answer
